\pkg{webchem}: An \proglang{R} Package to Retrieve Chemical Information from the Web

### 2020-03-21 revision3_2581/paper.tex
Updated code in the article

% citation alias for LUBW institution
\defcitealias{lubw_2016}{LUBW (2016)}
% \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\section[Implementation and design details]{Implementation and design details}
The \pkg{webchem} package is written entirely in \proglang{R} and
available under an MIT license.  The development repository is hosted
on \citet{github} and a stable version is released on the
Comprehensive \proglang{R} Archive Network (CRAN) and available at
\url{https://CRAN.R-project.org/package=webchem}.  \pkg{webchem} is
part of the rOpenSci project \citep{boettiger2015building}, which aims
at fully reproducible data analysis.

@@

Some data sources provide application programming interfaces (API).
Web APIs define functions that allow accessing services and data via
http and return data in a specific way.  \pkg{webchem} uses the API of
a data source provider, where available.  For sources where an API is
lacking, data is directly searched and extracted from the web pages,
analogous to manual interaction with a website.

@@

First we search for ETOX internal ID numbers using the substance names:
%
\begin{CodeChunk}
\begin{CodeInput}
R> subs <- unique(jagst$substance)
R> ids <- get_etoxid(subs, match = "best")
R> head(ids)
\end{CodeInput}
\begin{CodeOutput}
  etoxid                           match distance                  query
1   8932     2,4-Dimethylphenol ( 8932 )        0     2,4-Dimethylphenol
2   8494 4-Chlor-2-methylphenol ( 8494 )        0 4-Chlor-2-methylphenol
3   <NA>                            <NA>     <NA>     4-para-nonylphenol
4   8397                Atrazin ( 8397 )        0                Atrazin
5   7240                 Benzol ( 7240 )        0                 Benzol
6   7331        Desethylatrazin ( 7331 )        0        Desethylatrazin
\end{CodeOutput}
\end{CodeChunk}
%
Only three substances could not be found in ETOX.  Here we specify
that only the \code{"best"} match (in terms of the Levenshtein
distance between query and results) is returned.  A manual check
confirms appropriate matches.  Other options include: \code{"all"} --
returns all matches; \code{"first"} -- returns only the first match
(not necessarily the best match); \code{"ask"} -- this enters an
interactive mode, where the user is asked for a choice if multiple
matches are found and \code{"na"} which returns \code{NA} in case of
multiple matches.

@@

In the same manner, we can now query other identifiers from another
source using these CAS numbers (see Figure~\ref{fig:fig1}), like
PubChem.
%
\begin{CodeChunk}
\begin{CodeInput}
R> cids <- get_cid(etox_cas)
R> pc_data <- pc_prop(cids, properties = "CanonicalSMILES")
R> pc_smiles <- smiles(pc_data)
\end{CodeInput}
\end{CodeChunk}
%
We can use SMILES to query ChemSpiderIDs and subsequently convert them to InChiKeys.
%
\begin{CodeChunk}
\begin{CodeInput}
R> csids <- get_csid(pc_smiles, from = "smiles")$csid
R> cs_inchikey <- cs_convert(csids, from = "csid", to = "inchikey")
\end{CodeInput}
\end{CodeChunk}
%
Finally, we combine the queried data into one data frame
%
\begin{CodeChunk}
\begin{CodeInput}
R> res <- data.frame(name = subs, cas = etox_cas, smiles = pc_smiles, 
+    cid = pc_data$CID, inchikey = cs_inchikey, csid = cs_data$csid, 
+    stringsAsFactors = FALSE)
\end{CodeInput}
\end{CodeChunk}
%
Note that in order to use the ChemSpider functions, a personal
authentication key is needed, which can be retrieved
from the ChemSpider web page.  Finally, we obtain a compound table
containing many different identifiers (Table~\ref{tab:comptable}),
allowing easy identification and merging with other data sets, e.g.,
the \code{lc50} data set based on CAS.

@@

\subsection[Identify common names and roles of chemicals]{Identify common names and roles of chemicals}
\pkg{webchem} can query the ChEBI database \citep{chebi} to retrieve common names and roles of chemicals on biota. Therefore, we first order the \code{lc50} data set by the three most toxic (i.e. lowest value) chemicals and then use \code{get_chebiid()} to retrieve ChEBI identifiers and substance names. The function returns a list of tables containing the ChEBI identifier, the chemical name, a search score and entity stars for the found entry. The searchscore is calculated by ChEBI and represents how well the query string matches one of the database entries. The entitystar column reflects the quality of the ChEBI data, also provided by the database.

\begin{CodeChunk}
\begin{CodeInput}
R> lite <- get_chebiid(lc50_3$cas)
R> lite <- do.call(rbind, lite) # bind to data.frmae
R> lite$cas <- rownames(lite)
R> lite$chebiasciiname[4] <- "tefluthrin" # erronous result
\end{CodeInput}
\end{CodeChunk}

We can now use the ChEBI identifiers to query the complete entitiy by using \code{chebi_comp_entity()}. From the result, we extract the parents table and refine the data to chemical roles. Subsequently, we prepare the query result and merge the data with the \code{lite} object to include the CAS numbers (i.e. our initial query strings) to the result.

\begin{CodeChunk}
\begin{CodeInput}
R> comp <- chebi_comp_entity(lite$chebiid) # ChEBI ids
R> comp <- do.call(rbind, lapply(comp, `[[`, "parents")) # extract parents table
R> comp$chebiid <- gsub('\\.[0-9]+', '', rownames(comp))
R> comp <- comp[ comp$type == "has role", ] # refine to chemical roles
R> role <- aggregate(chebiName ~ chebiid,
                     data = comp,
                     FUN = function(x) paste0(x, collapse = ', '))
R> merge(lite[ names(lite) %in% c('cas', 'chebiid', 'chebiasciiname') ],
         role,
         by = 'chebiid')
\end{CodeInput}
\end{CodeChunk}

In doing so we retrieve Table~\ref{tab:chebi}, which provides information on the individual roles of the chemicals and we observe that, for example ethion, the most toxic pesticide towards \textit{D. magna} in our table is designed to be used as an acari- and insecticide, which inhibts the acetylcholinesterase enzyme.

\begin{table}[t!]
\centering
\begin{tabular}{p{2cm}p{2cm}p{8cm}}
  \hline
  cas & name & roles \\
  \hline
  563-12-2 & ethion & insecticide \newline environmental contaminant \newline EC 3.1.1.7 (acetylcholinesterase) inhibitor \newline acaricide \newline agrochemical \\
  96182-53-5 & tebupirimfos & EC 3.1.1.7 (acetylcholinesterase) inhibitor \\
  3383-96-8 & temephos & EC 3.1.1.7 (acetylcholinesterase) inhibitor \newline acaricide \newline agrochemical \newline ectoparasiticide \\
\hline
\end{tabular}
\caption{CAS numbers, names and chemical roles for the three most toxic substances towards \textit{D. magna} according to ChEBI.}
\label{tab:chebi}
\end{table}

>>>>

% citation alias for LUBW institution
\defcitealias{lubw_2016}{LUBW (2020)}
% \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\section[Implementation and design details]{Implementation and design details}
The \pkg{webchem} package is written entirely in \proglang{R} and
available under an MIT license.  The development repository is hosted
on GitHub \citep{github} and a stable version is released on the
Comprehensive \proglang{R} Archive Network (CRAN) and available at
\url{https://CRAN.R-project.org/package=webchem}.  \pkg{webchem} is
part of the rOpenSci project \citep{boettiger2015building}, which aims
at fully reproducible data analysis.

@@

Some data sources provide application programming interfaces (API).
Web APIs define functions that allow accessing services and data via
http and return data in a specific way.  \pkg{webchem} uses the API of
a data source provider, where available.  For sources where an API is
lacking, but policies of the service provider permit programmable access, data is directly searched and extracted from the web pages,
analogous to manual interaction with a website.

@@

First we search for ETOX internal ID numbers using the substance names:
%
\begin{CodeChunk}
\begin{CodeInput}
R> subs <- unique(jagst$substance)
R> ids <- get_etoxid(subs, match = "best")
R> head(ids)
\end{CodeInput}
\begin{CodeOutput}
  etoxid                           match distance                  query
1   8668            2,4-Xylenol ( 8668 )        0     2,4-Dimethylphenol
2   8494 4-Chlor-2-methylphenol ( 8494 )        0 4-Chlor-2-methylphenol
3   <NA>                            <NA>     <NA>     4-para-nonylphenol
4   8397                Atrazin ( 8397 )        0                Atrazin
5   7240                 Benzol ( 7240 )        0                 Benzol
6   7331        Desethylatrazin ( 7331 )        0        Desethylatrazin
\end{CodeOutput}
\end{CodeChunk}
%
Only three substances could not be found in ETOX.  Here we specify
that only the \code{"best"} match (in terms of the Levenshtein
distance between query and results) is returned.  A manual check
confirms appropriate matches.  Other options include: \code{"all"} --
returns all matches; \code{"first"} -- returns only the first match
(not necessarily the best match); \code{"ask"} -- this enters an
interactive mode, where the user is asked for a choice if multiple
matches are found and \code{"na"} which returns \code{NA} in case of
multiple matches.

@@

In the same manner, we can now query other identifiers from another
source using these CAS numbers (see Figure~\ref{fig:fig1}), like
PubChem.
%
\begin{CodeChunk}
\begin{CodeInput}
R> cids <- get_cid(etox_cas)
R> pc_data <- pc_prop(cids, properties = "CanonicalSMILES")
R> pc_smiles <- smiles(pc_data)
\end{CodeInput}
\end{CodeChunk}
%
We can use SMILES to query ChemSpiderIDs and subsequently convert them to InChiKeys.
%
\begin{CodeChunk}
\begin{CodeInput}
R> csids <- get_csid(pc_smiles, from = "smiles")
R> cs_inchikey <- cs_convert(csids$csid, from = "csid", to = "inchikey")
\end{CodeInput}
\end{CodeChunk}
%
Finally, we combine the queried data into one data frame
%
\begin{CodeChunk}
\begin{CodeInput}
R> res <- data.frame(name = subs, cas = etox_cas, smiles = pc_smiles, 
+    cid = pc_data$CID, inchikey = cs_inchikey, csid = csids$csid, 
+    stringsAsFactors = FALSE, row.names = NULL)
\end{CodeInput}
\end{CodeChunk}
%
Note that in order to use the ChemSpider functions, a personal
authentication key is needed, which can be retrieved
from the ChemSpider web page.  Finally, we obtain a compound table
containing many different identifiers (Table~\ref{tab:comptable}),
allowing easy identification and merging with other data sets, e.g.,
the \code{lc50} data set based on CAS.

@@

\subsection[Identify common names and roles of chemicals]{Identify common names and roles of chemicals}
We might be interested in common names and roles of the most toxic chemicals in the \code{lc50} data set, i.e., chemicals with the lowest $LC_{50}$ values. \pkg{webchem} can query the ChEBI database \citep{chebi} to retrieve common names and roles of chemicals on biota. We can extract the CAS numbers of the three most toxic chemicals, and then use \code{get_chebiid()} to retrieve ChEBI identifiers and substance names. We can then use the acquired ChEBI identifiers to query the complete entity by using \code{chebi_comp_entity()}. From the result, we can extract the parents table and refine the data to chemical roles.

\begin{CodeChunk}
\begin{CodeInput}
R> cas_rns <- lc50[order(lc50$value)[1:3],"cas"]
R> chebiids <- get_chebiid(cas_rns)
R> comp <- chebi_comp_entity(chebiids$chebiid)
R> pars <- lapply(comp, function(x) with(x, parents[parents$type == "has role",]))
\end{CodeInput}
\end{CodeChunk}

Finally we can compile the data into a table. Table~\ref{tab:chebi} provides information on the individual roles of the three most toxic chemicals and we observe that, for example ethion, the most toxic pesticide towards \textit{D. magna} in our table, is designed to be used as an acari- and insecticide, which inhibts the acetylcholinesterase enzyme.

\begin{table}[t!]
\centering
\begin{tabular}{p{2cm}p{2cm}p{8cm}}
  \hline
  cas & name & roles \\
  \hline
  563-12-2 & ethion & insecticide \newline environmental contaminant \newline EC 3.1.1.7 (acetylcholinesterase) inhibitor \newline acaricide \newline agrochemical \\
  96182-53-5 & tebupirimfos & EC 3.1.1.7 (acetylcholinesterase) inhibitor \\
  3383-96-8 & temephos & EC 3.1.1.7 (acetylcholinesterase) inhibitor \newline acaricide \newline agrochemical \newline ectoparasiticide \\
\hline
\end{tabular}
\caption{CAS numbers, names and chemical roles from ChEBI for the three most toxic substances in the \code{lc50} data set towards \textit{D. magna}.}
\label{tab:chebi}
\end{table}
